Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-04-19 paper/paper.tex

\section{Relation to existing cognitive accounts on default mode function}
\subsection{Predictive coding hypothesis}
Predictive coding mechanisms
\citep{clark2013whatever, friston2008hierarchical}
are a frequently evoked idea in the context of default mode function
\citep{bar2007}.
Cortical responses are explained as
emerging from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
Feed-forward sensory processing is constantly calibrated by
top-down modulation from more multi-modal and associative brain regions
further away from primary sensory cortical regions.
The dynamic interplay between cortical processing levels
may enable inferences about the world by reconciling
gaps between fresh sensory input and expectations computed
from stored prior information.
At each stage of neural processing,
an internally generated prediction of aspects of sensory sensation is
directly compared against the actual environmental input.
A prediction error at one of the processing levels
incurs plasticity changes of neuronal back projections
(i.e., adapting model parameters)
to allow for gradually improved future prediction of the environment.
In this way,
the predictive coding hypothesis offers explanations for
the constructive, generative nature of sensory perception
\citep{friston2010free} and
the intimate relation of motor action to sensory expectations
\citep{wolpert1995internal, kording2004bayesian}.
Contextual sensorimotor integration of perception-action cycles
may be maintained by top-down modulation
by a-priori information about the environment.

Predictive coding processes
permit updates of the internal representation of the environment
to best accommodate and prepare the organism
for the constant influx of sensory examples and
for reacting to the environment.
There are hence a number of common properties
between the predictive coding account
and the proposed formal account.
Importantly,
a generative model of how perceived sensory cues arise in the
world would be incorporated into
the current neuronal wiring.
Further,
both frameworks are plausibilized by neuroscientific evidence that suggest
the human brain to be a ``statistical organ'' with the biological purpose to
generalize from the past to new experiences.
Neuroanatomically, axonal
back projections indeed outnumber by far the axonal input projections
existing in the primate brain \citep{salin1995corticocortical}.
Moreover,
both accounts provide a parsimonious explanation why the
human brain decreases processing load devoted to incoming information
when the environment is predictable because the internal generative
model only requires updating following descrepancies between
environmental reality and its internal reinstantiation.
Increased computation resources are however allocated
when novel stimuli or
unexpected events are encountered by the organism.
The predictive coding and formal account
naturally evoke a mechanism of brain plasticity in that
neuronal wiring gets increasingly adapted
when faced by unprecedented environmental challenges.

While sensory experience is a generative process from both views,
the predictive coding account frames
sensory perception of the external world
as a generative experience due to the modulatory top-down influence at
various stages of sensory input processing.
In our RL view, instead,
the Markov property of our sequential decision making framework entails
the environmental model to be
\textbf{effectively expressed in the current state of the human agent
XXX guys, please double-check}.
Further,
the hierarchical processing aspect from predictive coding
is re-expressed in our account in form of
nested prediction of probable upcoming actions, states, and rewards.
Both frameworks model the consequences of action.
The predictive coding framework is typically explained without
explicit parameterization of the agent's horizon and
has a tendency to be presented as emphasizing prediction about the
immediate future.
In the present account, the horizon of that
look into the future is explicitly expressed in the $\gamma$ parameter
of the Bellman equation.
% \item \suggestremove{Mismatch negativity of PC is immanent in reward prediction error
% by the difference between actually predicted reward of an action given
% a state and the reward predicted by non-linear regression with graded
% discounting of future rewards.}
Finally,
the process of adapting the neuronal connections
for improved top-down modulation
takes the concrete form of gradient computation and
back-propagation in our RL implementation.
It is however important to note that
the neurobiological plausibility of
the back-propagation procedure is controversal
\citep{goodfellow2016deep}.
In sum,
the present formal account may serve as a concrete implementation of
the long-standing predictive coding account from cognitive neuroscience.
MPDs have the advantage of exposing an explicit
mechanism for the horizon of future considerations or
how the internal model of the world gets updated,
and
more explicitly reconciles sensory input processing and action consideration.
%
Recasting DMN function in terms of MDPs therefore naturally incorporates
a majority of aspects of the prediction coding hypothesis.

@@

In comparison,
both the semantic hypothesis and MDPs expose mechanisms of
how action considerations could be mentally explored.
In both accounts,
there is also no evidence to indicate that predictions of various
levels of complexity, abstraction, timescale and purpose
use mechanisms that are qualitatively different. This concurs with
DMN activity increases across time, space, and content domains
demonstrated in many neuroimaging studies \citep{spreng2009common}.
The semantic hypothesis
and MDP account provide explanations why hippocampus lesion does
not only impair recalling memories, but also hypothetical and future
thinking \citep{hassabis2007patients}.
While both semantic hypothesis and
our formal account propose memory-based internally
generated information for probabilistic mental models of action outcomes,
MDPs render explicit the grounds on which the final action is
eventually chosen, namely, the estimated cumulative reward.
\textbf{The notion of semantic knowledge associations is
incorporated into the MDP as the Markov property,
that is, the current state directly results from the
agent's history of states and actions. The learned
estimations of value and action transitions drive
stimulus processing and action choice in the present.}
The MDPs naturally integrate the former egocentric
(more related to current action, state and reward) and the later
allocentric (more related to past and future actions, states, and rewards)
angles on the world in a same optimization problem.
Further,
the semantic account of DMN function does not offer
a mechanistic explanation how explicit world knowledge and semantic analogies thereof
lead to prediction of future actions and states.
The semantic hypothesis does not explain why memory recall
for scene construction in humans is typically fragmentary and noisy
instead of accurate and reliable.
In contrast to existing accounts on semantics and
mental scene construction, the random and creative aspects of DMN function
are explained in MDPs by the advantages of stochastic optimization.
Yet, the MDP framework provides an algorithmic explanation in that
stochasticity of the parameter space explored
by the Monte Carlo solvers achieves better fine-tuning of the
action policies and estimation of expected reward outcomes.
That is, the purposeful stochasticity of policy and value estimation
in MDPs provides a candidate explanation for why humans
have evolved imperfect noisy memories
as the more advantageous adaptation.
In sum, episodic scene construction according to the semantic
account is lacking an explicit time and incentive model.

@@

\section{Conclusion}
What brain function could be important enough
for the existence and survival of the human species
to warrant constant, high energy costs?
MDPs provide an attractive
process model how the human association cortex
might implement supra-modal representation and control of the environment to
optimize the organism's fate.
% by obtaining reward and avoiding pain.
This idealized process model explains
a number of previous experimental observations in the
DMN by a simple but non-trivial mechanism.
%
From the computational view of a Markovian sequential decision process,
behavior unfolds by integrating happened past events
and possible future events to guide action choice in the present context.
This functional account is more compatible with the DMN's
poorly understood involvement across
autobiographical memory recall, problem solving,
abstract reasoning based on internally generated scenes, social cognition,
as well as delay discounting and self-related prospection.
MDPs provide a mathematical formalism how
optimal substructure in the environment can be recursively exploited
when confronted with complicated decisions.
Improvement of the internal world model
by injecting stochasticity in the recall of past
actions and outcomes may explain why
very accurate memories have been disfavored in human evolution
and why human creativity might be adaptive.
%
In principle,
neuroscientific experiments can be designed that operationalize
the set of action, value, and state variables that determine
the behavior of intelligent RL systems.
The proposed machine-learning perspective
on DMN biology is hence not only practically computable but
also yields falsifiable neuroscientific hypotheses.
At the least, we propose an alternative vocabulary to
describe and interpret experimental findings in neuroscience studies
on the DMN.
%
Ultimately,
the DMN can be viewed as a functional integrator
from the relatively recent events to anticipated upcoming events
in order to constantly improve action in our dynamic world.

>>>>

\section{Relation to existing cognitive accounts on default mode function}
\subsection{Predictive coding hypothesis}
Predictive coding mechanisms
\citep{clark2013whatever, friston2008hierarchical}
are a frequently evoked idea in the context of default mode function
\citep{bar2007}.
Cortical responses are explained as
emerging from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
Feed-forward sensory processing is constantly calibrated by
top-down modulation from more multi-modal and associative brain regions
further away from primary sensory cortical regions.
The dynamic interplay between cortical processing levels
may enable inferences about the world by reconciling
gaps between fresh sensory input and expectations computed
based on stored prior information.
At each stage of neural processing,
an internally generated prediction of aspects of environmental sensations is
directly compared against the actual environmental input.
A prediction error at one of the processing levels
incurs plasticity changes of neuronal back projections
(i.e., adapting model parameters)
to allow for gradually improved future prediction of the environment.
In this way,
the predictive coding hypothesis offers explanations for
the constructive, generative nature of sensory perception
\citep{friston2010free} and
the intimate relation of motor action to sensory expectations
\citep{wolpert1995internal, kording2004bayesian}.
Contextual sensorimotor integration of perception-action cycles
may be maintained by top-down modulation
using a-priori information about the environment.

In short, predictive coding processes
conceptualize updates of the internal representation of the environment
to best accommodate and prepare the organism
for the constant influx of sensory examples and
performing action on the environment.
There are hence a number of common properties
between the predictive coding account
and the proposed formal account based on MDPs.
Importantly,
a generative model of how perceived sensory cues arise in the
world would be incorporated into
the current neuronal wiring.
Further,
both accounts are plausibilized by neuroscientific evidence that suggest
the human brain to be a ``statistical organ'' with the biological purpose to
generalize from the past to new experiences.
Neuroanatomically, axonal
back projections indeed outnumber by far the axonal input projections
existing in the primate brain \citep{salin1995corticocortical}.
Moreover,
both accounts provide a parsimonious explanation why the
human brain decreases processing load devoted to incoming information
when the environment is predictable because the internal generative
model only requires updating following descrepancies between
environmental reality and its internal reinstantiation.
Increased computation resources are however allocated
when novel stimuli or
unexpected events are encountered by the organism.
The predictive coding and formal account
naturally evoke a mechanism of brain plasticity in that
neuronal wiring gets increasingly adapted
when faced by unprecedented environmental challenges.

While sensory experience is a generative process from both views,
the predictive coding account frames
sensory perception of the external world
as a generative experience due to the modulatory top-down influence at
various stages of sensory input processing.
In our RL view, instead,
the Markov property of our sequential decision making framework entails
the environmental model to be
\textbf{effectively expressed in the current state of the human agent
XXX guys, please double-check}.
Further,
the hierarchical processing aspect from predictive coding
is re-expressed in our account in form of
nested prediction of probable upcoming actions, states, and rewards.
Both accounts model the consequences of action.
The predictive coding account is typically explained without
explicit parameterization of the agent's horizon and
has a tendency to be presented as emphasizing prediction about the
immediate future.
In the present account, the horizon of that
look into the future is explicitly expressed in the $\gamma$ parameter
of the Bellman equation.
% \item \suggestremove{Mismatch negativity of PC is immanent in reward prediction error
% by the difference between actually predicted reward of an action given
% a state and the reward predicted by non-linear regression with graded
% discounting of future rewards.}
Finally,
the process of adapting the neuronal connections
for improved top-down modulation
takes the concrete form of gradient computation and
back-propagation in our RL implementation.
It is however important to note that
the neurobiological plausibility of
the back-propagation procedure is controversal
\citep{goodfellow2016deep}.
In sum,
the present formal account may serve as a concrete implementation of
the long-standing predictive coding account from cognitive neuroscience.
MPDs have the advantage of exposing an explicit
mechanism for the horizon of future considerations or
how the internal model of the world gets updated,
and
more explicitly reconciles sensory input processing and action consideration.
%
Recasting DMN function in terms of MDPs therefore naturally incorporates
a majority of aspects of the prediction coding hypothesis.

@@

In comparison,
both the semantic hypothesis and MDPs expose mechanisms of
how action considerations could be mentally explored.
In both accounts,
there is also no evidence to indicate that predictions of various
levels of complexity, abstraction, timescale and purpose
use mechanisms that are qualitatively different. This concurs with
DMN activity increases across time, space, and content domains
demonstrated in many neuroimaging studies \citep{spreng2009common}.
The semantic hypothesis
and MDP account provide explanations why hippocampus lesion does
not only impair recalling memories, but also hypothetical and future
thinking \citep{hassabis2007patients}.
While both semantic hypothesis and
our formal account propose memory-based internally
generated information for probabilistic mental models of action outcomes,
MDPs render explicit the grounds on which the final action is
eventually chosen, namely, the estimated cumulative reward.
\textbf{The notion of semantic knowledge associations is
incorporated into the MDP as the Markov property,
that is, the current state directly results from the
agent's history of states and actions. The learned
estimations of value and action transitions drive
stimulus processing and action choice in the present.}
The MDPs naturally integrate the former egocentric
(more related to current action, state and reward) and the later
allocentric (more related to past and future actions, states, and rewards)
angles on the world in a same optimization problem.
Further,
the semantic account of DMN function does not offer
a mechanistic explanation how explicit world knowledge and semantic analogies thereof
lead to prediction of future actions and states.
The semantic hypothesis does not explain why memory recall
for scene construction in humans is typically fragmentary and noisy
instead of accurate and reliable.
In contrast to existing accounts on semantics and
mental scene construction, the random and creative aspects of DMN function
are explained in MDPs by the advantages of stochastic optimization.
Yet, the MDP-based account provides an algorithmic explanation in that
stochasticity of the parameter space explored
by the Monte Carlo solvers achieves better fine-tuning of the
action policies and estimation of expected reward outcomes.
That is, the purposeful stochasticity of policy and value estimation
in MDPs provides a candidate explanation for why humans
have evolved imperfect noisy memories
as the more advantageous adaptation.
In sum, episodic scene construction according to the semantic
account is lacking an explicit time and incentive model.

@@

\section{Conclusion}
What brain function could be important enough
for the existence and survival of the human species
to warrant constantly high energy costs?
MDPs provide an attractive
formal account how the human association cortex
might implement supra-modal representation and control of the environment to
optimize the organism's fate.
% by obtaining reward and avoiding pain.
This idealized process model explains
a number of previous experimental observations in the
DMN by a simple but non-trivial mechanism.
%
From the computational view of a Markovian sequential decision process,
behavior unfolds by integrating happened past events
and possible future events to guide action choice in the present context.
MDPs provide a mathematical formalism how
optimal substructure in the environment can be recursively exploited
when confronted with complicated decisions.
%
This functional interpretation is more compatible with the DMN's
poorly understood involvement across
autobiographical memory recall, problem solving,
abstract reasoning based on internally generated scenes, social cognition,
as well as delay discounting and self-related prospection.
Improvement of the internal world model
by injecting stochasticity in the recall of past
actions and outcomes may explain why
very accurate memories have been disfavored in human evolution
and why human creativity might be adaptive.
%
In principle,
neuroscience experiments can be designed that operationalize
the set of action, value, and state variables that determine
the behavior of intelligent RL systems.

The proposed machine-learning perspective
on DMN biology is hence not only practically computable but
also yields falsifiable neuroscientific hypotheses.
%
At the least, we propose an alternative vocabulary to
describe and interpret experimental findings in neuroscience studies
on this macroscopic brain network.
%
Ultimately,
the DMN can be viewed as a functional integrator
from the relatively recent events to anticipated upcoming events
in to constantly refine action in our dynamic world.
